---
title: "Introduction to R with Tidyverse"
author: "Sophie Lee"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: TRUE
  html_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 5: R packages and add-ons

R packages are a collection of functions and datasets developed by R users that expand existing base R capabilities or add completely new ones. Packages allow users to apply the most up-to-date methods shortly after they are developed, unlike other statistical software packages that require an entirely new version. The quickest way to install a package in R is:

```install.packages(‘name of package’)```

R requires an internet connection to install a package from R's online repository, CRAN. Packages must be downloaded before their first use on a machine and to install any updates that have been made since the last installation. 

Once a package has been installed, use the following code to load it into your R session:

```require(name of package)```

or

```library(name of package)```

Packages must be reloaded every time a new session of R is opened. Therefore, it is good practice to include the ```require``` or ```library``` command at the beginning of a script file. 

All available packages are listed on the [CRAN website](https://cran.r-project.org/web/packages/available_packages_by_name.html) with details of what each one does. Use the ```installed.packages( )``` for a list of all installed packages on your machine.

## Chapter 6: Introduction to the Tidyverse and data wrangling

### 6.1 Tidyverse

Up to this point in the course, we have been using a style of R coding language (or syntax) known as base R, the original version of the language that does not require any additional packages. However, there are different approaches that have been developed within R packages. For example, the Tidyverse set of packages, which aim to improve certain aspects of R syntax. 

The choice of syntax style depends on personal preference. Base R is very stable as it is not dependent on packages which may be updated regularly. However, the Tidyverse set of packages have been developed to make R code more readable and flexible than base R. We will be using a combination of base R and Tidyverse coding style within this course to utilise the extensive data manipulation tools developed for Tidyverse. 

To load tidyverse packages into your R session, run the following code:

```{r Install tidyverse, eval = FALSE}
install.packages(‘tidyverse') 
```

```{r Attach tidyverse, warning = FALSE}
library(tidyverse) 
```


### 6.2 The working directory
The working directory is a file path on your computer that R sets as the default location when opening, saving, or exporting documents, files, and graphics. It is possible to specify this location manually but setting the working directory saves time and makes code more efficient. 

To set the working directory, choose *Session -> Set Working Directory -> Change Directory…* from the drop-down menu and select the location you want to act as the directory. These steps must be taken every time R is restarted, otherwise it will be reset to R’s default (usually ‘My Documents’). This can be done using the ```setwd``` function: 

```{r Set working directory, eval = FALSE}
setwd("C://My Documents/Training")
```

Sets the working directory to the ‘Training’ folder in ‘My Documents’.

```{r Get working directory, eval = FALSE}
get(wd)
```

Returns the current working directory

```{r List files, eval = FALSE}
list.files( )
```

Returns a list of all the files that currently exist within your working directory. This is particularly useful when loading datasets from the directory into R as they can be copy and pasted into the script to avoid spelling mistakes and error messages.

#### 6.2.1 RStudio projects

Although working directories can be set manually using the ```setwd``` command, this requires an absolute file path which makes it liable to break easily (for example, if the folder is moved to a different location) and makes it difficult to share code with others. An alternative approach is to use RStudio projects which keeps all associated files, including data, script files and outputs, grouped together. Rather than requiring an absolute file path, *projects* (files with the extension *.Rproj*) set the working directory relative to their location. 

To create a new project, choose *File -> New project…* from the drop-down menu. The project can either be created in a brand-new directory, where a working directory does not already exist, or within an existing directory. Existing projects can be opened by selecting *File -> Open project…* from the drop-down menu. 

RStudio projects help create an efficient workflow by grouping all files necessary for each analysis project together and assists collaboration by setting a working directory relative to the .Rproj files’ location. 

### 6.3 Data input

The easiest way to upload data from excel is to convert it into a comma-delimited file (*.csv*). R has functions that can easily read and convert csv files into objects within the R environment. Before converting files, it is essential to check that they are correctly formatted. Excel files should only contain one sheet of data with no pictures or graphics, each row should correspond to a case or observation and each column should correspond to a variable. 

Once the spreadsheet has been converted into a .csv file, it can be read into R using the ```read_csv``` function. For example, the following code reads in the core spending power report from 2020 from *data* folder of the current working directory and saves it as an object named ```csp_2020```:

```{r Read CSP 2020 data}
csp_2020 <- read_csv("data/CSP_2020.csv")
```

If the file is saved somewhere other than the current working directory, the file path would need to be included in the brackets. 

This function assumes that the first row of the .csv file contains variable names. If that is not the case, the argument ```col_names = FALSE``` must be added after a comma in the function. 

For full details of the default arguments used in ```read_csv```, search for the help file using the ```?read_csv``` command.

Imported datasets will appear in the Environment window of the console once they are saved as objects. This window shows the number of variables and observations in each object. To view the contents of an object, click on its name in the Environment window or use the function ```View(data)```. Useful functions that help explore a dataset include:

```{r Variable names}
names(csp_2020)
```

Returns a list of variable names

```{r Structure}
str(csp_2020)
```

Displays information about the structure of an object. This differs depending on the type of object entered. For example, this object is a tibble (Tidyverse's name for a dataset). The information given about tibbles includes the object dimensions (396 x 9, or 396 rows and 9 columns), variable names, and variable types.

```{r Head function}
head(csp_2020)
```

Returns the first 6 rows of the tibble.

```{r Tail function}
tail(csp_2020)
```

Returns the final 5 rows of the tibble.

### 6.4 Selecting variables
There are a number of ways to select individual variables within a dataset within base R or tidyverse.

To select one or more variables using tidyverse, we can use the ```select``` function:

```select(data, variable1, variable2, ...)```

For example, if we wanted to return the new homes bonus (nhb) for each local authority, the seventh column of the dataset, we can either select this based on the variable name or its location in the object:

```{r Selecting variables}
select(csp_2020, nhb_2020)

select(csp_2020, 7)
```

We can select multiple variables and return them as a tibble by separating the variable names or numbers with commas:

```{r Selecting multiple variables}
select(csp_2020, ons_code, authority, region)
```

When selecting consecutive variables, a shortcut can be used that gives the first and last variable in the list, separated by a colon, ```:```. The previous example can be carried out using the following code:

```{r Selecting multiple with colon}
select(csp_2020, ons_code:region)
```

The ```select``` function can be combined with a number of other ‘helper’ functions that allow us to select variables with certain characteristics or naming conventions. For example:

```starts_with("xyz")``` returns all variables where the name begins "xyz".

```ends_with("xyz")``` returns all variables where the name ends "xyz".

```everything()``` returns all variables.

For a full list of selection helpers, checkthe help file ```?tidyr_tidy_select```. 

The ```select``` function returns the variable(s) in the form of a tibble (or dataset). However, some functions, such as basic summary functions, require data to be entered as a vector (a list of values). Tibbles with a single variable can be converted into a vector using the ```as.vector``` function, or we can use the base R approach to selecting a single variable which returns the data as a vector by default: 

```{r Base R selecting, eval = F}
csp_2020$nhb_2020
```

It is important to save any changes made to the existing dataset. This can be done using the ```write_csv``` function:

```{r Write CSV files, eval = FALSE}
write_csv(csp_2020, file = "data/csp_2020_new.csv")
```

When saving updated data, be sure to give it a different name to the raw data file, otherwise the original data will be overwritten. It is important to keep a copy of the raw data.


### 6.5 Filtering data

Filtering data allows us to select subgroups based on conditional statements. If the observation meets the conditional statement, it will be returned. The ```filter``` function from the Tidyverse package ```dplyr``` can be used to return these subgroups:

```filter(data, condition 1, condition 2, ...)```

Conditional statements include less than ```<```, less than or equal to ```<=```, greater  than ```>```, greater or equal to ```>=```, is equal to ```==```, is not equal to ```!=```, or is missing ```is.na(variable)```. 

For example, to return the core spending power for local authorities in the North West region:

```{r Filter NW}
filter(csp_2020, region == "NW")
```

We can use multiple conditional statements to filter a subgroup, these statements must be separated in the function using commas. For example, to return a subgroup of local authorities that were in the North West region and had a settlement funding assessment (SFA) of over £40 million, we use the following:

```{r Filter NW with SFA > 40}
filter(csp_2020, region == "NW", sfa_2020 > 40)
```

### 6.6 Pipes

The ```filter``` and ```select``` functions can be combined to create an analysis dataset that just contains data we need for our current problem. Rather than carrying out these functions separately and creating multiple objects, we can use a 'pipe'. A pipe is represented be the ```%>%``` symbol and can be read as 'and then' within code. Pipes allow us to carry out multiple steps in the same process whilst ensuring the code is still readable. For example, we can return a list of local authority names from the North West region:

```{r Pipe NW authority names}
csp_2020 %>% 
  filter(region == "NW") %>% 
  select(authority)
```

The pipe can also be combined with the ```filter``` function to ```count``` the number of observations that lie within a subgroup:

```{r Count NW region}
csp_2020 %>% 
  filter(region == "NW") %>% 
  count()
```

### 6.7 Adding new variables

In Tidyverse, the function ```mutate``` allows us to add new variables, either by manually specifying them or by creating them from existing variables. Multiple variables can be created within the same function and are separated by a comma. For example, we can create a new variables with the squared settlement funding assessment, and another that recodes the council tax variable into a categorical variable with three levels (low: below £5 million, medium: between £5 million and £15 million, and high: above £15 million):

```{r Mutate function}
csp_2020_new <- csp_2020 %>% 
  mutate(sfa_2020_sq = sfa_2020 ^ 2,
         ct_2020_cat = cut(ct_total_2020, breaks = c(0, 5, 15, Inf),
                           labels = c("Low", "Medium", "High"),
                           include_lowest = TRUE))
```

### Exercise 3

1. How many local authorities were included in the London region?

2. Give three ways that it would be possible to select all spend variables (sfa_2020, nhb_2020, etc.) from the CSP_2020 dataset.

3. Create a new tibble, ```em_2020```, that just includes local authorities from the East Midlands.
  - How many local authorities in the East Midlands had an SFA of between £5 and 10 million? (**Hint**: check the ```filter``` help file for a useful helper)
  - Create a new variable with the total overall spend in 2020 for local authorities in the East Midlands.

