---
title: "Introduction to R with Tidyverse"
author: "Sophie Lee"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: TRUE
  html_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 5: R packages and add-ons

R packages are a collection of functions and datasets developed by R users that expand existing base R capabilities or add completely new ones. Packages allow users to apply the most up-to-date methods shortly after they are developed unlike other statistical software packages that require an entirely new version. The quickest way to install a package in R is:

```install.packages(‘name of package’)```

Once a package has been installed, use the following code to load it into your R session:

```require(name of package)```

or

```library(name of package)```

R requires an internet connection to install a package. After installation, packages must be loaded every time R is opened. It is good practice to include the require or library command at the beginning of a script file. All available packages listed on the [R website](https://cran.r-project.org/web/packages/available_packages_by_name.html) with details of what each one does. Use the ```installed.packages( )``` for a list of all installed packages on your machine.

## Chapter 6: Opening and exploring data

### 6.1 Tidyverse

Up to this point in the course, we have been using a style of R coding language (or syntax) known as base R, the original version of the language and one that does not require any additional packages. However, there are different approaches that have been developed within R packages. For example, the Tidyverse set of packages, which aim to improve certain aspects of R syntax. 

The choice of syntax style depends on personal preference. Base R is very stable as it is not dependent on packages which may be updated regularly. However, the Tidyverse set of packages have been developed to make R code more readable and flexible than base R. We will be using a combination of base R and Tidyverse coding style within this course to utilise the extensive data manipulation tools developed for Tidyverse. 

To load tidyverse packages into your R session, run the following code:

```{r Install tidyverse, eval = FALSE}
install.packages(‘tidyverse') 
```

```{r Attach tidyverse}
library(tidyverse) 
```


### 6.2 The working directory
The working directory is a file path on your computer that R sets as the default location when opening, saving, or exporting documents, files, and graphics. It is possible to specify this location manually but setting the working directory saves time and makes code more efficient. 

To set the working directory, choose *Session -> Set Working Directory -> Change Directory…* from the drop-down menu and select the location you want to act as the directory. These steps must be taken every time R is restarted, otherwise it will be reset to R’s default (usually ‘My Documents’). This can be done using the ```setwd``` function: 

```{r Set working directory, eval = FALSE}
setwd("C://My Documents/Training")
```

Sets the working directory to the ‘Training’ folder in ‘My Documents’.

```{r Get working directory, eval = FALSE}
get(wd)
```

Returns the current working directory

```{r List files, eval = FALSE}
list.files( )
```

Returns a list of all the files that currently exist within your working directory. This is particularly useful when loading datasets from the directory into R as they can be copy and pasted into the script to avoid spelling mistakes and error messages.

#### 6.2.1 RStudio projects

Although working directories can be set manually using the ```setwd``` command, this requires an absolute file path which makes it liable to break easily (for example, if the folder is moved to a different location) and makes it difficult to share code with others. An alternative approach is the use of RStudio projects which keeps all associated files, including data, script files and outputs, grouped together. Rather than requiring an absolute file path, *projects* (files with the extension *.Rproj*) set the working directory relative to their location. 

To create a new project, choose *File -> New project…* from the drop-down menu. The project can either be created in a brand-new directory, where a working directory does not already exist, or within an existing directory. Existing projects can be opened by selecting *File -> Open project…* from the drop-down menu. 

RStudio projects help create an efficient workflow by grouping all files necessary for each analysis project together and assists collaboration by setting a working directory relative to the .Rproj files’ location. 




